Data Analysis and Modeling of Teen Vaccination Surveys

### 2024-04-05 code/analysis-code/analysis3.qmd
Random decision tree and improving LASSO
RF tree is in analysis 2.5 document and LASSO improved at bottom of analysis3 document. Maybe cross validation tuning and refitting, like we did in class, would make this model the best. RF tree did not turn out well, but I did not use it to make predictions yet.

# Remove rows with missing values
mydata$SEX <- droplevels(mydata$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$INS_STAT2_I <- droplevels(mydata$INS_STAT2_I, exclude = "MISSING Data")
mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
```
Next, we will split the data into training and testing groups so that the LASSO regression has a comparison group.
```{r}
# Split data into training and testing datasets
set.seed(123) # seed for reproducibility
split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
train_data <- training(split_data)
test_data <- testing(split_data)
```
Next, we will specify the LASSO recipe, or the regression formula. This will contain the predictors of interest used for previous analysis and `P_UTDHPV` as the outcome. 
```{r}
# Create recipe containing predictors of interest
lasso_rec <- recipe( P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes())  %>% # convert categorical variables (nominal) into dummy variables
  step_rm(all_outcomes(), -all_outcomes()) #remove the outcome variables from the dataset

@@

```{r}
# Specify logistic regression model
lasso_spec <- logistic_reg(penalty = 0.005, mixture = 1) %>%
  set_engine("glmnet")

@@

```{r}
# Generate predictions
predictions <- lasso_fit %>%
  predict(new_data = mydata) %>%
  bind_cols(mydata)

# Calculate ROC AUC
predictions %>%
  mutate(predicted_numeric = as.numeric(.pred_class)) %>%
  roc_auc(P_UTDHPV, predicted_numeric)

# Generate predictions
predictions2 <- lasso_fit %>%
  predict(new_data = train_data) %>%
  bind_cols(train_data)

# Calculate ROC AUC
predictions2 %>%
  mutate(predicted_numeric = as.numeric(.pred_class)) %>%
  roc_auc(P_UTDHPV, predicted_numeric)

```
When attempting to determine how lowering the penalty effected the ability of the model, we produced some predictions with both the original data set and training data, both performed fairly poorly, with a ROC_AUC score close to 0.5, suggesting the the model created predictions that were comparable to random.

@@

>>>>

# Remove rows with missing values
mydata$SEX <- droplevels(mydata$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$INS_STAT2_I <- droplevels(mydata$INS_STAT2_I, exclude = "MISSING Data")
mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
```
Next, we will split the data into training and testing groups so that the LASSO regression has a comparison group.
```{r}
# Split data into training and testing datasets
set.seed(123) # seed for reproducibility
split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
train_data <- training(split_data)
test_data <- testing(split_data)
```
Next, we will specify the LASSO recipe, or the regression formula. This will contain the predictors of interest used for previous analysis and `P_UTDHPV` as the outcome. 
```{r}
# Create recipe containing predictors of interest
lasso_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes())  %>% # convert categorical variables (nominal) into dummy variables
  step_rm(all_outcomes(), -all_outcomes()) #remove the outcome variables from the dataset

@@

```{r}
# Specify logistic regression model
lasso_spec2 <- logistic_reg(penalty = 0.005, mixture = 1) %>%
  set_engine("glmnet")

# Fit the model using training data
lasso_fit2 <- lasso_wf %>%
  add_model(lasso_spec2) %>%
  fit(data = train_data)

# Extract and tidy model parameters
lasso_fit2 %>%
  extract_fit_parsnip() %>%
  tidy()

# Extract and tidy model parameters
lasso_coef <- lasso_fit2 %>%
  extract_fit_parsnip() %>%
  tidy()

@@

```{r}
# Make predictions for the first LASSO model
predictions <- lasso_fit %>%
  augment(new_data = test_data)

# Calculate ROC AUC
roc_auc <- predictions %>%
  roc_auc(truth = P_UTDHPV, .pred_UTD)
print(roc_auc)

# Make predictions for the LASSO model with lower penalty
predictions2 <- lasso_fit2 %>%
  augment(new_data = test_data)

# Calculate ROC AUC
roc_auc <- predictions2 %>%
  roc_auc(truth = P_UTDHPV, .pred_UTD)
print(roc_auc)

```
When attempting to determine how lowering the penalty effected the ability of the model, we produced some predictions for the model with a higher penalty and the model with a very low penalty. Lowering the penalty from 0.05 to 0.005 improved the ROC AUC to from 0.5 to 0.37. This suggests that the second model is more informative. However, to improve the model further, we should find the ideal penalty parameter.

@@

(**Rachel redundantly added code that does the same thing. We can choose our favorite of the two next time we meet.**)
I will now tune the penalty parameter using cross validation to find the penalty with the best roc_auc.
```{r}
lasso_penalty <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

folds <- rsample::vfold_cv(train_data, v = 5)
tune_grid <- tibble(penalty = 10^seq(-2, -1, length.out = 10))

wf <- workflow() %>%
  add_recipe(lasso_rec)%>%
  add_model(lasso_penalty)

tune_wf <- wf %>% 
  tune_grid(resamples = folds,
            grid = tune_grid,
            control = control_grid(verbose = FALSE, save_pred = TRUE),
            metrics = metric_set(roc_auc))

# Select the best model based on ROC AUC
best_mod <- tune_wf %>% 
  select_best(metric = "roc_auc")

final_wf <- finalize_workflow(wf, best_mod)
print(final_wf)
```

# LASSO Re-fit
Now, we will run the LASSO regression again, but with the tuned penalty parameter of 0.01. 
```{r}
# Specify logistic regression model
lasso_spec3 <- logistic_reg(penalty = 0.01, mixture = 1) %>%
  set_engine("glmnet")

# Fit the model using training data
lasso_fit3 <- lasso_wf %>%
  add_model(lasso_spec3) %>%
  fit(data = train_data)

# Extract and tidy model parameters
lasso_fit3 %>%
  extract_fit_parsnip() %>%
  tidy()

# Extract and tidy model parameters
lasso_coef <- lasso_fit3 %>%
  extract_fit_parsnip() %>%
  tidy()

# Make predictions for ROC_AUC with the tuned model
predictions3 <- lasso_fit3 %>%
  augment(new_data = test_data)

# Calculate ROC AUC
roc_auc <- predictions3 %>%
  roc_auc(truth = P_UTDHPV, .pred_UTD)
print(roc_auc)
```
The ROC_AUC for this model with a penalty parameter of 0.01 is now 0.39. This is slightly worse than the model with a penalty of 0.05. However, the model with a lower penalty is more "unstable", so doing cross validation for the ROC_AUC may reveal which model is truly ideal.

# Cross Validation
Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.01 to determine which model performs best when presented with "new data".
